\documentclass[
	%a4paper, % Use A4 paper size
	letterpaper, % Use US letter paper size
]{jdf}

\usepackage[linesnumbered, ruled]{algorithm2e}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{amsthm}

\addbibresource{references.bib}

\author{Ashish Narasimham, Michael Walker, Sakae Watanabe}
\email{}
\title{CS7643: Group Project Proposal}


\begin{document}

\maketitle

\section*{Team Name}

Alfalfa Hunters

\section*{Project Title}

I Like the Stock - Applying Modern Supervised Approaches for Algorithmic Trading

\section*{Project Summary}
The goal of this project is to develop a system that produces actionable trading signals (buy, hold, or sell) for financial instruments based on historical time-series data. By leveraging various deep learning techniques, we aim to accurately predict target trades, facilitating the creation of a profitable investment strategy. This system is of particular interest to users who are looking to optimize their investment portfolios and enhance their financial security.

Existing research is inherently limited, as every published algorithm is a direct loss of alpha for the publisher. We hope to develop an intuitive understanding of the algorithmic trading space, or at least a subset of the vast approaches, in order to understand how machine learning can be applied to a structured space with many competing agents. We’ll apply recent techniques to learn how the state of the art performs


\section*{Approach}

\begin{itemize}
    \item We will attempt 3 different approaches (corresponding to the 3 team members) to understand what is successful - liquid neural networks \cite{hasani_liquid_2020}, spatio-temporal transformers \cite{boyle_spatiotemporal_2023}, and some form of RNN. We will implement the techniques as described in the respective papers \citep{kazemi_time2vec_2019,su_self-attentive_2022}.
    \item We will do feature engineering, hyperparameter tuning, try to bring in supporting datasets (other financial instruments that may provide predictive signal for our objective)
    \item We will describe each technical approach in the report, supported by mathematical analysis
\end{itemize}

\section*{Resources/Related Work \& Papers}

\printbibliography[heading=none]

\section*{Datasets(Provide a link to the dataset)}

\begin{enumerate}
    \item We will use Yahoo Finance data since it is readily available at the Daily granularity across just about every stock symbol for a ~3 year historical timeframe
    \item \$SPY buy-and-hold will be our benchmark
    \item We will focus on 5 high-volatility small cap healthcare stocks:
    \begin{enumerate}
        \item \$BIVI - BioVie, Inc.
        \item \$VTAK - Catheter Precision, Inc.
        \item \$CYCC - Cyclacel Pharmaceuticals, Inc.
        \item \$ATNF - 180 Life Sciences Corp.
        \item \$BIAF - bioAffinity Technologies, Inc.
    \end{enumerate}
\end{enumerate}


\section*{Group Members}

\begin{itemize}
	\item Ashish Narasimham
	\item Michael Walker
	\item Sakae Watanabe
\end{itemize}

\end{document}
